
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ACKNOWLEDGMENT %%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Acknowledgement}

{
\small
\emph{First and foremost, to my beloved wife, Ilena - this journey has been yours as mine.}
\vspace{0.6cm}

Five years ago, relocating to Sweden for a Ph.D. journey seemed unimaginable. At that time, I was grappling with the decision of whether to advance my academic pursuits. During this phase of uncertainty, my mentor and friend, Oscar Luis Vera, presented a life-changing opportunity: a Ph.D. position in Software Engineering at KTH. Now, as I look back over these past five years, I see a journey marked by challenges. Each step, no matter how difficult, has contributed immeasurably to my growth and learning. I carry with me a deep sense of accomplishment and no regrets—my heartfelt thanks to Oscar and Ali.

I want to thank the distinguished members of my jury. Professor Sukyoung Ryu, thanks for accepting to take the role of opponent in my defense. Professor Dilian Gurov, thanks for taking the advance review of this work. Professors, Quentin Stiévenart, Bjorn De Sutter, and Weihang Wang, thank you all for being part of my grading committee and for your invaluable insights.

I would like to extend my gratitude to my supervisors, Benoit Baudry and Martin Monperrus. Their guidance has been a constant, ensuring that I never felt lost. Their support extended beyond professional advice, blossoming into a genuine friendship. There were times when I might have been challenging to work with, but I deeply appreciate your patience and willingness to listen - all that many histories about Cuba could be boring :) - Thank you, Benoit and Martin.

Over the past five years, a significant period in my life, I have had the privilege of meeting and working with many people. Each one has played a unique role in contributing to my work. I extend my heartfelt gratitude to my lab colleagues for their unwavering support and collaboration. Special thanks to Cesar Soto, Nicolas Harrand, Romi Tsoupidi, Javier Ron, Erik Gustavsson, Nadia Campo, Orestis Floros, Long Zhang, He Ye, Deepika Tiwari, Zimin Chen, Khashayar Etemadi, Amir Ahmadian,  Mikhail Shcherbakov, Yi Liu, André Silva and Anoud Alshanak for their invaluable contributions and camaraderie.

These last five years have been supported by the Stiftelsen för Strategisk Forskning(SSF) and the Trustfull project. I would like to thank all my colleagues in this project, especially Professor Musard Balliu for his invaluable insights.

To my Cuban friends in Sweden. Our roots in Cuba instill a deep sense of community and connection, often accompanied by an emotional intensity that makes each recognition meaningful. Understanding this shared sentiment, I choose to embrace all of you collectively in this paragraph without writing names. It's a way to ensure that no one feels overlooked. This is my heartfelt acknowledgment to all of you, my friends.

Last but never least, to my family in Cuba, especially to “mi abuela Mary”. Family values led me to this successful journey.
}